Miniproject - Algebraic Geometry

### 2021-12-10 Algebraic geometry/submission.tex
Added graphs and statement about the largest independent set in the graphs to alg geo

\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{stmaryrd}
\usepackage[style=english]{csquotes}
\usepackage[language=english, backend=biber, style=alphabetic, sorting=nyt]{biblatex}

@@

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered co-lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{colex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    Note that for $i < j < u < v$ have $x_{u, v}, x_{j, v} \succ x_{i, v} \succ x_{i, j}, x_{i, u}, x_{j, u}$.
    Thus the leading term of $x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}$ is
    \begin{equation*}
        \mathrm{lt}(x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}) = x_{i, v} x_{j, u}
    \end{equation*}
    It follows that $J \subseteq \mathrm{lt}(I)$.

@@

\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 4$.
\end{lemma}
\printbibliography
\end{document}

>>>>

\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{stmaryrd}
\usepackage[style=english]{csquotes}
\usepackage[language=english, backend=biber, style=alphabetic, sorting=nyt]{biblatex}
\usepackage{tikz}

@@

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered co-lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{colex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    Note that for $i < j < u < v$ have $x_{u, v}, x_{j, v} \succ x_{i, v} \succ x_{i, j}, x_{i, u}, x_{j, u}$.
    Thus the leading term of $x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}$ is
    \begin{equation*}
        \mathrm{lt}(x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}) = x_{i, v} x_{j, u}
    \end{equation*}
    It follows that $J \subseteq \mathrm{lt}(I)$.

@@

\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
Now we want to study how the independent sets in $G_n$ look like.
\begin{example}
    \label{ex:Gn}
    The graph $G_4$ is the following
    \\\includegraphics[width = \textwidth]{graph4.pdf}
    The graph $G_5$ is the following
    \\\includegraphics[width = \textwidth]{graph5.pdf}
    The graph $G_6$ is the following
    \\\includegraphics[width = \textwidth]{graph6.pdf}
\end{example}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 4$.
\end{lemma}
\begin{proof}
    The idea is to write the graph in layers (or as a kind of ``pyramid'') as displayed in Example~\ref{ex:Gn}.
    Then we can ``push'' an independent set onto the bottom layer, by repeatedly taking vertices in the set that are maximally high up and on the outside, and replacing them by their ``child''. 
    This shows the claim, as the bottom layer (plus the 4 unconnected vertices) has size $2n - 4$.
\end{proof}
\printbibliography
\end{document}
